Hypertension and microbiome

### 2019-10-14 articletwo.Rmd
Salt plot

```{r libraries, cache = FALSE}
library(dplyr)
library(tibble)
library(phyloseq)
library(nortest)
library(microbiome)
library(knitr)
library(tidyr)
library(vegan)
library(reshape)
library(parallel)
library(officer)
library(flextable)
library(xtable)
library(rvg)
library(tableone)
library(scales)
library(ggplot2)
library(gridExtra)
library(png)
library(pander)
library(ggpubr)
library(broom)
library(ggfortify)
library(RColorBrewer)
library(gvlma)
library(purrr)
library(pwr)
library(gtable)
library(car)
library(M3C)
library(emmeans)
library(DESeq2)
```

@@

    dds <- lapply(c2l(var.BP), function(x) {
        dds.data <- phyloseq_to_deseq2(pseq.genus.core.deseq,
                                       as.formula(sprintf("~ %s + %s",
                                                          paste0(var.CL, collapse = "+"), x)))
        DESeq(dds.data,
              test="Wald",
              fitType="parametric",
                     parallel = TRUE,
                     BPPARAM=MulticoreParam(16))})

@@

```{r deseq continuous}
dset.deseq.signf <- deseqresults(dds, names.dset)
```

@@

```{r lacto data frame}
df.genus.salt <- pseq.genus %>%
    microbiome::transform(transform = "clr") %>%
    prune_taxa(c("g_Lactobacillus"), .) %>%
    psmelt %>%
    dplyr::filter(!is.na(NA.)) %>%
    spread(OTU, Abundance) %>%
    dplyr::mutate(dUNA = myscale(NA.),
                  salt_groups = relevel(factor(dplyr::ntile(dUNA, 3)), ref='1'),
                  lacto_groups = relevel(factor(dplyr::ntile(g_Lactobacillus, 3)), ref='1'))
```

Lactobacillus prevalence

```{r lactobacillus prevalence}
pseq.genus %>%
    microbiome::transform(transform = "compositional") %>%
    prune_taxa(c("g_Lactobacillus"), .) %>%
    psmelt %>%
    spread(OTU, Abundance) %>%
    dplyr::mutate(lacto.prevalent = ifelse(g_Lactobacillus > 0.001, TRUE, FALSE),
                  duna.present = ifelse(is.na(NA.), 0, 1)) %>%
    group_by(duna.present) %>%
    summarize(lacto_detected = sum(lacto.prevalent), n = n(), ratio = lacto_detected/n) %>%
    kable
  ```

Characteristics in salt analysis. Analyses are limited by the number
of participants with collected dU-NA.

```{r salt characteristics}
df.genus.salt %>%
    select(-Sample) %>% 
    gather(key,value) %>%
    group_by(key) %>%
    summarise_all(funs(N = n(),
                       Mean = mean(as.numeric(value), na.rm = TRUE),
                       ones = sum(value == 1, na.rm = TRUE),
                       sd = sd(as.numeric(value), na.rm = TRUE),
                       NAs = sum(is.na(value))))  %>%
    mutate_if(is.numeric, round, 3) %>%
    kable()
```

Required sub sample size

```{r required sodium subsample size}
lapply(c2l(seq(1, 1.7, 0.05)), function(or) 
    powerMediation::powerLogisticCon(n = nrow(df.genus.salt),
                                     p1 = 0.4,
                                     OR = or)) %>%
    map_df(., ~as.data.frame(.x), .id = "or") %>%
    kable
```

### How sodium affects lactobacillus abundances?

Associations between continuous dU-NA and lactobacillus abundances.

```{r abund salt lm continuous}
lm.lacto.duna <- lm(g_Lactobacillus ~ dUNA + BL_AGE + SEX, data = df.genus.salt)
lm.lacto.duna %>% tidy %>% mutate_if(is.numeric, round, 3) %>% kable
```

Testing assumptions for linear regression analysis

```{r abund salt lm continuous checks table}
gvlma.table.plot(lm.lacto.duna)$table %>% kable
```

Diagnostic plots for linear model

```{r abund salt lm continuous checks plot, fig.width=12, fig.height=14}
gvlma.table.plot(lm.lacto.duna)$plot %>% plot
```

### Is there associations between salt groups and abudances?

Associations between categorical dU-NA split in three equal size groups and lactobacillus abundances

```{r abund salt lm group}
lm.lacto.saltgroup <- lm(g_Lactobacillus ~ salt_groups + BL_AGE + SEX, data = df.genus.salt)
lm.lacto.saltgroup %>% tidy %>% mutate_if(is.numeric, round, 3) %>% kable
```

Testing assumptions for linear regression analysis

```{r abund salt lm group checks table}
gvlma.table.plot(lm.lacto.saltgroup)$table %>% kable
```

Diagnostic plots for linear model

```{r abund salt lm group checks plot, fig.width=12, fig.height=14}
gvlma.table.plot(lm.lacto.saltgroup)$plot %>% plot
```

Analysing also differences between salt groups using ANOVA.

```{r salt anova}
aov.lacto.saltgroup <- aov(g_Lactobacillus ~ salt_groups + BL_AGE + SEX, data = df.genus.salt)
aov.lacto.saltgroup %>% xtable %>% kable
```

Levene's test for equality of variance

```{r salt levene}
car::leveneTest(g_Lactobacillus ~ salt_groups, data = df.genus.salt) %>% kable
```

Tukey's range for pairwise test between salt groups

```{r salt HSD}
hsd.lacto.saltgroup <- TukeyHSD(aov.lacto.saltgroup, "salt_groups")
hsd.lacto.saltgroup$salt_groups %>% kable
```

### Adjusting means for forest plot

Calculate adjusted means for lactobacillus assuming male sex and mean age.

```{r lacto adjusted means}
adjmean.lacto.saltgroup <- emmeans::emmeans(lm.lacto.saltgroup, specs = c("salt_groups")) %>%
    as.data.frame %>%
    dplyr::rename(term = salt_groups, estimate = emmean, conf.low = lower.CL, conf.high = upper.CL)
```

Adjusted means

```{r forest lims}
adjmean.lacto.saltgroup %>% kable
```

## Is there associations between lactobacillus abudance and hypertension?

Studying associations between blood pressure variables and lactobacillus abundances

```{r lactobacillus RR}
calculateglm(df.genus.salt,
             modelstring = "%s ~ g_Lactobacillus + BL_AGE + SEX",
             filterstr = "Lactobacillus") %>% 
    select(response, term, estimate, std.error, p.value) %>%
    mutate_if(is.numeric, round, 3) %>%
    kable
```

Checking assumptions for linear regression for MAP

```{r table map lactobacillus}
gvlma.table.plot(lm(MAP ~ g_Lactobacillus + BL_AGE + SEX, data = df.genus.salt))$table %>% kable
```

Diagnostic plots for linear model

```{r plot map lactobacillus, fig.width=12, fig.height=14}
gvlma.table.plot(lm(MAP ~ g_Lactobacillus + BL_AGE + SEX, data = df.genus.salt))$plot %>% plot
```

Odds for hypertension by categorical lactobacillus abundance.

```{r glm htn lacto groups}
glm.hypertension.lactogroup <- glm(HYPERTENSION ~ lacto_groups + BL_AGE + SEX,
                  family = binomial(link=logit),
                  data = df.genus.salt)

odds.hypertension.lactogroup <- glm.hypertension.lactogroup %>%
    broom::tidy(conf.int = TRUE, conf.level = 0.95, exponentiate = TRUE) %>%
    filter(!grepl("Intercept", term)) %>%
    select(term, estimate, conf.low, conf.high) %>%
    add_row(term = "lacto_groups1", estimate = 1.0, conf.low = NA, conf.high = NA) %>%
    dplyr::filter(grepl("lacto", term))

glm.hypertension.lactogroup %>%
    broom::tidy(conf.int = TRUE, conf.level = 0.95, exponentiate = TRUE) %>%
    mutate_if(is.numeric, round, 3) %>%
    kable
```

Odds for hypertension

```{r odds htn lacto}
odds.hypertension.lactogroup <- glm.hypertension.lactogroup %>%
    broom::tidy(conf.int = TRUE, conf.level = 0.95, exponentiate = TRUE) %>%
    filter(!grepl("Intercept", term)) %>%
    select(term, estimate, p.value, conf.low, conf.high) %>%
    add_row(term = "lacto_groups1", estimate = 1.0, conf.low = NA, conf.high = NA) %>%
    dplyr::filter(grepl("lacto", term))
odds.hypertension.lactogroup %>% kable
```

Power analysis of odds of hypertension

```{r power hypertention}
power.htn <- pwr.f2.test(u = 3, v=NULL, f2 = 0.02, sig.level=0.05, power=0.8)
round(power.htn$v + power.htn$u + 1)

pwr.f2.test(u = 3, v=836, f2 = NULL, sig.level=0.05, power=0.95)
```

## Twin-forest plot

```{r forest plot lacto}
gsalt <- grid.arrange(
    myforestplot(adjmean.lacto.saltgroup,
                 xlab = "Tertile of 24-hour urinary sodium",
                 ylab = "Lactobacillus abundance",
                 subtitle = "A.",
                 my_y_scale = scale_y_continuous(limits = c(5.4, 5.65), breaks = seq(0, 10, 0.1))),
    myforestplot(odds.hypertension.lactogroup,
                 xlab = "Tertile of Lactobacillus abundance",
                 ylab = "Odds of hypertension",
                 subtitle = "B.",
                 my_y_scale = scale_y_log10(limits = c(0.4, 1.5), breaks = seq(0.5, 2.0, 0.5)),
                 interceptone = TRUE),
    ncol=2)

ggsave(file = "cache/salt-forest.png", plot=gsalt, height=3, width=5.5, dpi = 300)
```

```{r Write figure 4, include=FALSE}
fig4head <- "Associations between dietary salt, lactobacillus abundance, and odds for hypertension."
fig4foot <- "In panel A we see lactobacillus abundances in different dietary salt groups. The mean value is adjusted using age and sex. In panel B we split participants in four groups by lactobacillus abundances and calculate odds ratio for hypertension. Age and sex are included in logistic model. A 95 % CI is drawn in both panels."
writeimage(doc, 4, "cache/salt-forest.png", fig4head, fig4foot)
```

Testing salt model using maaslin

```{r salt model maaslin}
taxa(pseq.genus.core) %>% mygrep(word = "lacto")

pseq.genus.salt <- prune_taxa(c("g_Lactobacillus"), pseq.genus.core)

pseq.genus.salt.meta <- meta(pseq.genus.salt) %>%
    tibble::rownames_to_column(var = "rowname") %>%
    dplyr::filter(!is.na(NA.)) %>%
    dplyr::mutate(dUNA = myscale(NA.),
                  salt_groups = relevel(factor(dplyr::ntile(dUNA, 3)), ref='1')) %>%
    tibble::remove_rownames() %>%
    tibble::column_to_rownames(var = "rowname") %>%
    select(var.BP, var.CL.min, starts_with("salt_groups"))

phyloseq::sample_data(pseq.genus.salt) <- phyloseq::sample_data(pseq.genus.salt.meta)

maaslin.salt <- maaslinwrapper(pseq = pseq.genus.salt,
                               looped = var.BP,
                               forced = c(var.CL.min, "salt_groups"),
                               taxa = "g_Lactobacillus")

maaslin.salt %>%
    merge(names.dset %>% select(Covariate, Category, Name), by.x ="Variable", by.y = "Covariate") %>%
    dplyr::mutate(Qval = p.adjust(P.value, method="BH"),
                  Feature = gsub("[gs]_", "", Feature),
                  beta = Coefficient) %>%
    arrange(Feature)
```

```{r squared clinical, eval = FALSE, include = FALSE}
mylmrsqr <- function(df, response, loopcovariates) {
    model.list <- list()
    for (i in seq(length(loopcovariates))) {
        fo <- sprintf("%s ~ %s", response, paste(loopcovariates, collapse = " + "))
        message(sprintf("formula : %s", fo))
        model <- lm(as.formula(fo), data = df)
        model$call <- as.formula(fo)
        model.list[[i]] <- model
        loopcovariates <- loopcovariates[-length(loopcovariates)]
    }
    return(model.list)
}

af.models <- mylmrsqr(df = meta(pseq.species),
         response = "MAP",
         loopcovariates = c("BL_AGE", "SEX", "BMI", "CURR_SMOKE", "Q57X", "PREVAL_DIAB",
                            "ANYDRUG"))

af <- do.call(anova, af.models)

af %>%
    as.data.frame %>%
        mutate(Rsqr =  `Sum of Sq`/sum(`Sum of Sq`, na.rm = T)*100)

c(summary(lm(SYSTM ~ BL_AGE + SEX, meta(pseq.species)))$adj.r.squared,
summary(lm(DIASM ~ BL_AGE + SEX, meta(pseq.species)))$adj.r.squared,
summary(lm(PULSEPRESSURE ~ BL_AGE + SEX, meta(pseq.species)))$adj.r.squared,
summary(lm(MAP ~ BL_AGE + SEX, meta(pseq.species)))$adj.r.squared) %>% mean

c(summary(lm(SYSTM ~ BL_AGE + SEX + BMI, meta(pseq.species)))$adj.r.squared,
summary(lm(DIASM ~ BL_AGE + SEX + BMI, meta(pseq.species)))$adj.r.squared,
summary(lm(PULSEPRESSURE ~ BL_AGE + SEX + BMI, meta(pseq.species)))$adj.r.squared,
summary(lm(MAP ~ BL_AGE + SEX + BMI, meta(pseq.species)))$adj.r.squared) %>% sort

>>>>

```{r libraries, cache = FALSE}
library(dplyr)
library(tibble)
library(phyloseq)
library(nortest)
library(microbiome)
library(knitr)
library(tidyr)
library(vegan)
library(reshape)
library(parallel)
library(officer)
library(flextable)
library(xtable)
library(rvg)
library(tableone)
library(scales)
library(ggplot2)
library(gridExtra)
library(png)
library(pander)
library(ggpubr)
library(broom)
library(ggfortify)
library(RColorBrewer)
library(gvlma)
library(purrr)
library(pwr)
library(gtable)
library(car)
library(M3C)
library(emmeans)
library(DESeq2)
library(ggeffects)
```

@@

    lapply(c2l(var.BP), function(x) {
        dds.data <- phyloseq_to_deseq2(pseq.genus.core.deseq, deseq.formula(x, var.CL))
        DESeq(dds.data,
              test="Wald",
              fitType="parametric",
              parallel = TRUE,
              BPPARAM=MulticoreParam(16))})

@@

```{r deseq continuous}
dset.deseq.signf <- deseqresults(dds, names.dset)
dset.deseq.signf %>% kable
```

@@

```{r salt phyloseq object}
rownames.salt <- pseq.genus %>% meta %>% subset(!is.na(NA.)) %>% rownames
pseq.genus.salt <- pseq.genus %>%
    prune_samples(rownames.salt, .) %>%
    prune_taxa(taxa(pseq.genus.core), .)
dset.genus.salt <- prune_lactobacillus(pseq.genus.salt, transform = "clr")
```

```{r salt characteristics}
dset.genus.salt %>%
    mytableone() %>%
    kable
```

```{r lactobacillus prevalence}
dset.genus.salt %>%
    summarize(lacto_detected = sum(lacto.prevalent),
              n = n(),
              ratio = lacto_detected/n) %>%
    kable
```

```{r lacto scatter plot}
g.salt <- saltplot(dset.genus.salt)
ggsave(file = "cache/gsalt.png", plot=g.salt, height=4, width=4, dpi = 300)
```

```{r Write figure 4, include=FALSE}
fig4head <- "Associations between dietary salt, lactobacillus abundance, and odds for hypertension."
fig4foot <- "In panel A we see lactobacillus abundances in different dietary salt groups. The mean value is adjusted using age and sex. In panel B we split participants in four groups by lactobacillus abundances and calculate odds ratio for hypertension. Age and sex are included in logistic model. A 95 % CI is drawn in both panels."
writeimage(doc, 4, "cache/gsalt.png", fig4head, fig4foot)
```

Deseq for lactobacillus

```{r lacto deseq}
dds.data.salt <- phyloseq_to_deseq2(pseq.genus.salt, ~ BL_AGE + SEX + NA.)

dds.salt <- DESeq(dds.data.salt, test="Wald", fitType="parametric")

results(dds.salt, name = "NA.") %>%
    as.data.frame %>%
    tibble::rownames_to_column("Feature") %>%
    dplyr::mutate(qval = p.adjust(pvalue, method="BH")) %>%
    filter(Feature %in% "g_Lactobacillus") %>%
    kable
```

# Supplements
